\textbf{Parallel dragonfly algorithm} \\[2ex] \large High Performance Computing for Data Science course 2024/2025

### 2025-02-13 text-paper/paper.tex
inserted openMP description and started MPI section into report

\section{Performace tests}

The serial implementation seen above can be easily tried out using some test function 
with known global minima in order to check the overall performances. We need to 
report time execution for each test function, using the input dimension as independent variable;
since however we have multiple input parameters to compare, we also have to decide which one will 
be considered during the performance evaluation.

\begin{table}[H]
\begin{tabular}{| c | c | c | c |}
  \hline
  \textit{Nr of iterations} & \textbf{sphere [ms]} & \textbf{rosenblock [ms]} & \textbf{rastrigin [ms]}\\
  \hline
  10 \\
  \hline

  100 \\
  \hline

  1000 \\
  \hline

  10000 \\ 
  \hline

  100000 \\
  \hline
\end{tabular}

\end{table}

\begin{table}[H]
  \begin{tabular}{| c | c | c | c |}
    \hline
    \textit{Nr of iterations} & \textbf{sphere [ms]} & \textbf{rosenblock [ms]} & \textbf{rastrigin [ms]}\\
    \hline
    10 \\
    \hline

    100 \\
    \hline

    1000 \\
    \hline

    10000 \\ 
    \hline

    100000 \\
    \hline
  \end{tabular}
\end{table}

\noindent As we can see the execution time increases exponentially with the consequentially increase of
the input.

\section{Parallel implementation}

\section{Parameters Retrievement}

\section{Speedup and final conclusions}

>>>>

\section{parallelization with MPI}

OpenMPI is a library that enables code execution on multiple cores, wheter they belong
to the same CPU or not; it can be useful in case of performace improvement of those algorihms
that require to solve complex tasks, just like the dragonfly algorithm: in fact this library can be easily 
adapted to the illustrated implementation in order to boost performance.

\subsection*{MPI on chunks}
As we said earlier chunks are independent from each otherm but usually they need to comunicate
with their neighbours in order to keep updated the global status of the program execution. 
It is possible to run chunks groups simoultaneously, using MPI builtin function to connect 
the processes.  The instructions computed by each process are available in a custom function
called \textit{dragonfly-compute}; after calculating the ID of the initial and final chunks
and initializing message structures using the \textbf{offsetof} macro
the function will execute the update loop that calculates the new velocities and positions 
of the dragonfly for each chunk associated. To do this we first need to update the current
location of enemies and food present in the chunk using the function \textbf{computation\_status\_merge},
using the position of food and enemies of other chunks retrieved by the function \textbf{message\_broadcast};
finally the process will proceed to update current position and velocities of dragonfiles using 
the function \textbf{dragonfly\_compute\_step} illustrated earlier

\subsection*{MPI and parameters}
Since MPI has been useful in the parallelization of the chunks then it can also become useful in
the training of parameters

\section{OpenMP implementation}

OpenMP is a low level API that allows us to execute a sequence of instructions on different threads belonging
to the same CPU core, improving in this way the performances of our implementation. This feature can be used to 
speed up the execution of some slow functions, in our case these will be dragonfly-compute-step and message-accumulate

\subsection*{General use}

First of all it is necessary to understand how OpenMP should be inserted in the code: first of all we have to use 
a compiler that supports this technology, in our case we have gcc, which allows OpenMP execution thanks to the flag -fopenmp.
Second we need to include a pragma directive in our code, which will wrap the set of instructions that we are interested to execute
in parallel. Since the implemented results will perform loops mainly then it is fundamental to calculate the work balance between the number of iterations, if the main
body is a loop,
and the number of threads available. Each thread has to define the starting and the end point of the loop that has to execute,
calculating the starting index as $rank*ratio$ and the end one as $ratio*(rank+1)$, where $rank$ is the ID of the thread

\subsection*{dragonfly-compute-step}

One of the two functions that we are interested to execute using OpenMP is \textit{dragonfly-compute-step}:
if we take a look on the serial version we notice that each computation step is independent
from other iterations. This means that the execution flow can be split in many subtasks assigned
to threads. The following code snippet illustrate how all the procedure has been implemented
\begin{lstlisting}[language=C,caption={parallelized dragonfly-compute-step}]

void dragonfly_compute_step(Dragonfly *d, 
                            float *average_speed,
                            float *cumulated_pos, 
                            float *food, 
                            float *enemy, 
                            unsigned int N, 
                            unsigned int NR_THREADS) {

  unsigned int base_random = rand_r(&d->seed);
  unsigned int rest = d->N % NR_THREADS;
  unsigned int ratio = d->N / NR_THREADS;

  ...

  #pragma omp parallel num_threads(NR_THREADS)
    {

      unsigned int rank = omp_get_thread_num();
      //printf("rank=%d\n", rank);
      unsigned int base = ratio * rank;
      unsigned int limit = ratio * (rank + 1);
      inner_dragonfly_step(d, average_speed, 
                          cumulated_pos, food, 
                          enemy, N,
                          base, limit, base_random);
    }

  ...

  if (rest != 0) {
      unsigned int r_base = ratio * NR_THREADS;
      unsigned int r_end = d->N;

      inner_dragonfly_step(d, average_speed, 
                           cumulated_pos, 
                           food, enemy, N,
                           r_base, r_end, base_random);
    }
  }

  weights_step(&d->w);
}

\end{lstlisting}

\noindent After we got all basic informations the \textit{inner\_dragonfly\_step}
function can be executed, wrapping it into the \textbf{pragma} directive that we have seen earlier: in this way each thread will 
perform the necessary operations only on a specific portion of the overall positions and speeds arrays.
If the number of threads does not divide perfectly the size of the arrays then we compute the rest
returned from the ratio calculation and the remaining cells will be processed using the serial approach.

\subsection*{computation\_accumulate}

The \textit{computation\_accumulate}  must compute
the cumulated positions and speeds of the chunk and find the best food and enemy positions.
The implementation of a parallelized approach in this case  is more complicated to achieve, 
because threads should read and write on the same memeory areas which could be 
difficult to execute independently from each other. This time it is necessary to define
local buffers for each thread, so they can store temporary results without the risk of using
overlapping memeory locations. The parallel section provides that each thread must compute the local
cumulated sums of velocities and positions, then they must select the position with the best fitness value
in order to find the best local position of food and enemies. Once the local execution is completed
it will update the global status of the chunk using a critical section, with one thread that can execute that
specific portion of the program

\begin{figure}[H]
  \includegraphics[scale=0.2]{reduce-example.png}
  \centering
  \caption{Illustration of the reduce process performed by the computation\_accumulate function}
\end{figure}
